Variance reduction for MCMC methods via martingale representations

### 2020-01-20 aap/Revision/vr-mcmc-revision.tex
update

\def\rmd{\mathrm{d}}
\def\rme{\mathrm{e}}
\def\rset{\mathbb{R}}
\def\NtrainPath{T}
\def\TrainSet{\mathcal{T}}
\modulolinenumbers[5]
\newcommand{\proofendsign}{$\Box$}
\newtheorem{thm}{Theorem}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{remark}{Remark}
\newtheorem{cor}{Corollary}
\newtheorem{example}{Example}
\newenvironment{proof}{{\noindent \bf Proof }}
 {{\hspace*{\fill}\proofendsign\par\bigskip}}
\def\TV{\operatorname{TV}}
\newcommand{\jac}[1]{\operatorname{J} _{#1}}

@@

\section{Bounds for moments of  ULA}\label{sec:appendix}
For $\gamma>0$, the Markov kernel $R_\gamma$
on $(\mathbb R^d,\mathcal B(\mathbb R^d))$
is defined by
\begin{equation}
\label{eq:kernel}
R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \nabla U(x)\right\|^2\right\}dy,\quad A\in \mathcal B(\mathbb R^d),
\end{equation}
while, for $k,l\in\mathbb N$, $k\le l$,
the kernel $Q_{\boldsymbol{\gamma}}^{k,l}$ is
$
Q_{\boldsymbol{\gamma}}^{k,l}=R_{\gamma_l}\cdots R_{\gamma_k}.
$
 We make the following assumptions.

@@

Due to \cite[Lemma~14]{brosse:2018:diffusion}, conditions {\bf (H1)} and {\bf (H2)} imply that there exist $K_1 > 0, m > 0$, such that for any $x \not\in B_{K_1}(0)$ it holds
\begin{equation}
\label{eq:H3}
\langle \nabla U(x), x \rangle \geq \frac{m}{2}\|x\|^2
\end{equation}

\begin{lem} Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$. Then the kernel $R_\gamma$ from \ref{eq:kernel} satisfies drift condition \ref{eq:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_1^2}{8} + 2d + m$ with $K_1$ from \ref{eq:H3}, $m$ from {\bf (H2)}. 
\end{lem}

@@

$$
 \leq (1-\gamma m)\|x\|^2 + 2\gamma^2 L^2 \|x\|^2
$$
Thus, plugging everything into expression for $R_{\gamma}V$ and using $\gamma < \frac{m}{4L}$, we obtain
$$
R_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + 2\gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma (2d + m)
$$
Now let $x \in B(0,K_1)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
$$
R_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^d\right) \leq 
$$

$$
\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(\frac{25K_1^2}{8} + 2d + m)
$$
for some absolute constant $C_1$.
\end{proof}

Now the geometric ergodicity result readily follows from \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018}, namely, the following lemma holds:

\begin{lem} Let the potential $U(x), x \in \mathbb{R}^D$ satisfy conditions {\bf H1} and {\bf H3}. Then for $0 < \gamma < \overline{\gamma}=\frac{m}{4L^2}$, for any $x \in X$ it holds
$$
d_V(\delta_xQ^n_\gamma,\pi_\gamma) \leq C\rho^n\left(V(x) + \pi_\gamma(V)\right)
$$ 
with $V(x) = 1 + \|x\|^2$ and constants
$$
b = \gamma(3K_1^2 + 2D + C_1); \quad \overline{b} = b\exp^{-\frac{m\gamma}{2}} + d; \quad \varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right); \quad
$$

$$
C = \left(1 + \exp^{-\frac{m\gamma}{2}}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp^{-\frac{m\gamma}{2}} - \frac{2b}{1+d})}\right);
$$

$$
\rho = \exp^{-\gamma\left(\frac{m}{2} - 2\frac{3K_1^2 + 2D + C_1}{d}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp^{-\frac{m\gamma}{2}} + \frac{2b}{d+1})}}}
$$
\end{lem}

\begin{proof} Note that the condition {\bf H1} implies that the Markov kernel $Q^n_\gamma$ satisfies $(1,\varepsilon)$-Doeblin condition with $\varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right)$. Together with drift condition {\bf H2} it allows to apply \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018} with appropriate constants. 
\end{proof}

\begin{thm}
\label{exp_mom_ula}
Suppose that
\begin{eqnarray*}
\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty.
\end{eqnarray*}
Under assumptions {\bf (H1)} and {\bf (H2)}, for any bounded function $g$ on $\mathbb{R}^d$ 
we have
\begin{eqnarray*}
\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}g(X_{p})-\bar Q_l(X_l)\right)^{4}\right] \lesssim \frac{1}{\Gamma^4_{N+2,N+n+1}}
\end{eqnarray*}
 and
\begin{eqnarray*}
\bar Q_l(X_l)=\mathsf{E}\left[\left.\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}\right].
\end{eqnarray*}
Moreover, 
\begin{eqnarray*}
\mathsf{E}\left[\exp\left(\rho\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right]\leq C
\end{eqnarray*}
for some small enough $\rho>0,$  all natural $0<l\leq p<\infty$ and constant $C$ not depending on $l,p.$
\end{thm}
\begin{proof}
 Note that 
\[
\mathsf{E}\left[e^{-s\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})}\right]=\mathsf{E}\left[e^{-s\left(\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})-\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx\right)}\right]e^{-s\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx}
\]
 with $\Gamma_{l,j}=\sum_{k=l}^{j}\gamma_{k}.$ According to \cite[Theorem~3]{bms:exp_conc_note}, it holds
\[
\mathsf{E}\left[e^{-s\left(\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})-\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx\right)}\right]\leq e^{s\varkappa\sum_{k=l}^{j}\gamma_{k}^{2}}
\]
 for some $\varkappa>0.$ Since $\sum_{k=1}^{\infty}\gamma_{k}^{2}\leq\Gamma^{\star},$we
have 
\[
\mathsf{E}\left[e^{-s\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})}\right]\leq e^{s\varkappa\Gamma^{\star}}e^{-s\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx}.
\]
Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R}),$ we have
\[
\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=2\int\mu^{2}(x)\pi(x)\,dx>0
\]
and as a result 
\[
\zeta_{l,j}(s)\leq e^{s\varkappa\Gamma^{\star}}e^{-2s\Gamma_{l,j}\int\mu^{2}(x)\pi(x)\,dx}
\]
 for any $l\leq j.$
\end{proof}

>>>>

\def\rmd{\mathrm{d}}
\def\rme{\mathrm{e}}
\def\rset{\mathbb{R}}
\def\NtrainPath{T}
\def\TrainSet{\mathcal{T}}
\modulolinenumbers[5]
\newcommand{\proofendsign}{$\Box$}
\newtheorem{thm}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{remark}{Remark}
\newtheorem{cor}{Corollary}
\newtheorem{example}{Example}
\newenvironment{proof}{{\noindent \bf Proof }}
 {{\hspace*{\fill}\proofendsign\par\bigskip}}
\def\TV{\operatorname{TV}}
\newcommand{\jac}[1]{\operatorname{J} _{#1}}

%added command for indicator - SS
\newcommand{\indiacc}[1]{\mathbbm{1}_{\{#1\}}}

@@

\section{Bounds for moments of  ULA}\label{sec:appendix}
In this appendix we give some definitions and prove auxiliary results which are the used to bound the exponential moment of the additive functional of ULA chain (Theorem~\ref{th:exp_mom_ula}).
\par
For $\gamma>0$, the Markov kernel $R_\gamma$ on $(\mathbb R^d,\mathcal B(\mathbb R^d))$, corresponding to the Unadjusted Langevin Algorithm with step $\gamma$ is defined by
\begin{equation}
\label{eq:kernel}
R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \nabla U(x)\right\|^2\right\}dy,\quad A\in \mathcal B(\mathbb R^d),
\end{equation}
while, for $k,l\in\mathbb N$, $k\le l$,
the kernel $Q_{\boldsymbol{\gamma}}^{k,l}$ is
$
Q_{\boldsymbol{\gamma}}^{k,l}=R_{\gamma_l}\cdots R_{\gamma_k}.
$
We make the following assumptions.

@@

Due to \cite[Lemma~14]{brosse:2018:diffusion}, conditions {\bf (H1)} and {\bf (H2)} implies that there exist constants $K_1 > 0, m > 0$, such that for any $x \not\in B_{K_1}(0)$ it holds
\begin{equation}
\label{eq:H3}
\langle \nabla U(x), x \rangle \geq \frac{m}{2}\|x\|^2
\end{equation}

We would show that conditions {\bf (H1)} and {\bf (H2)} are sufficient to show that ULA kernel $R_\gamma$ satisfies the so-called drift condition:
\begin{definition}
\label{def:drift}
We say that the Markov kernel $R_\gamma$ satisfies Foster-Lyapunov drift condition if there exist a measurable function $V: X \rightarrow [1; +\infty)$, real numbers $\lambda \in (0,1)$ and $C > 0$ such that for any $x \in X$,
\begin{eqnarray*}
RV(x) \leq \lambda^{\gamma} V(x) + \gamma C
\end{eqnarray*}
\end{definition}

\begin{lem} Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$. Then the kernel $R_\gamma$ from \ref{eq:kernel} satisfies drift condition \ref{def:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_1^2}{8} + 2d + m$ where $K_1$ is from \ref{eq:H3}, $m$ is from {\bf (H2)}. 
\end{lem}

@@

$$
 \leq (1-\gamma m)\|x\|^2 + 2\gamma^2 L^2 \|x\|^2
$$
Thus, plugging everything into expression for $R_{\gamma}V$ and using $\gamma < \frac{m}{4L}$, we obtain
$$
R_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + 2\gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma (2d + m)
$$
Now let $x \in B(0,K_1)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
$$
R_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^2\right) \leq 
$$

$$
\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(\frac{25K_1^2}{8} + 2d + m)
$$
\end{proof}

For now let us assume for simplicity that we run ULA with fixed step size $\gamma$. Then it is known that under assumption {\bf H1} corresponding Markov chain would have unique stationary distribution $\pi_\gamma$, which is not equal to $\pi$. Yet this chain will be $V-$geometrically ergodic due to \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018}. Namely, the following lemma holds:

\begin{lem} Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf H1} and {\bf H2}. Then for $0 < \gamma < \overline{\gamma}=\frac{m}{4L^2}$, for any $x \in X$ it holds
$$
d_V(\delta_xQ^n_\gamma,\pi_\gamma) \leq C\rho^n\left(V(x) + \pi_\gamma(V)\right)
$$ 
with $V(x) = 1 + \|x\|^2$ and constants
$$
C = \left(1 + \exp^{-\frac{m\gamma}{2}}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp^{-\frac{m\gamma}{2}} - \frac{2b}{1+d})}\right);
$$

$$
b = \gamma(\frac{25K_1^2}{8} + 2d + m); \quad \overline{b} = b\exp^{-\frac{m\gamma}{2}} + d; \quad \varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right); \quad
$$

$$
\rho = \exp^{-\gamma\left(\frac{m}{2} - 2\frac{\frac{25K_1^2}{8} + 2d + m}{d}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp^{-\frac{m\gamma}{2}} + \frac{2b}{d+1})}}}
$$
\end{lem}

\begin{proof} Note that the condition {\bf (H1)} implies that the Markov kernel $Q^n_\gamma$ satisfies $(1,\varepsilon)$-Doeblin condition with $\varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right)$. Together with drift condition \ref{eq:drift} it allows to apply \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018} with appropriate constants. 
\end{proof}

\begin{thm}
\label{th:exp_mom_ula}
Suppose that
\begin{eqnarray*}
\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty.
\end{eqnarray*}
Under assumptions {\bf (H1)} and {\bf (H2)}, for any bounded function $g$ on $\mathbb{R}^d$ 
we have
\begin{eqnarray*}
\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}g(X_{p})-\bar Q_l(X_l)\right)^{4}\right] \lesssim \frac{1}{\Gamma^4_{N+2,N+n+1}}
\end{eqnarray*}
 and
\begin{eqnarray*}
\bar Q_l(X_l)=\mathsf{E}\left[\left.\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}\right].
\end{eqnarray*}
Moreover, 
\begin{eqnarray*}
\mathsf{E}\left[\exp\left(\rho\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right]\leq C
\end{eqnarray*}
for some small enough $\rho>0,$  all natural $0<l\leq p<\infty$ and constant $C$ not depending on $l,p.$
\end{thm}
\begin{proof}
 Note that 
\[
\mathsf{E}\left[e^{-s\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})}\right]=\mathsf{E}\left[e^{-s\left(\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})-\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx\right)}\right]e^{-s\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx}
\]
 with $\Gamma_{l,j}=\sum_{k=l}^{j}\gamma_{k}.$ According to \cite[Theorem~3]{bms:exp_conc_note}, it holds
\[
\mathsf{E}\left[e^{-s\left(\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})-\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx\right)}\right]\leq e^{s\varkappa\sum_{k=l}^{j}\gamma_{k}^{2}}
\]
 for some $\varkappa>0.$ Since $\sum_{k=1}^{\infty}\gamma_{k}^{2}\leq\Gamma^{\star},$we
have 
\[
\mathsf{E}\left[e^{-s\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})}\right]\leq e^{s\varkappa\Gamma^{\star}}e^{-s\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx}.
\]
Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R}),$ we have
\[
\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=2\int\mu^{2}(x)\pi(x)\,dx>0
\]
and as a result 
\[
\zeta_{l,j}(s)\leq e^{s\varkappa\Gamma^{\star}}e^{-2s\Gamma_{l,j}\int\mu^{2}(x)\pi(x)\,dx}
\]
 for any $l\leq j.$
\end{proof}
